To my colleagues and collaborators in the DEVOLAB and BEACON, I benefited greatly from your insight and your camaraderie.
Thank you.
Notable mentions here include Dr. Acacia Ackles, Dr. Wolfgang Banzhaf, Cliff Bohm, Dr. Emily Dolson, Austin Ferguson, Jose Hernandez, Dr. Alex Lalejini, Dr. Josh Nahum, Dr. Anselmo Pontes, and Kate Skocelas, and Dr. Anya Vostinar.

Thank you to my mentees for your valuable work.
Sara Boyd and Tait Wecht made huge improvements to the Empirical library's web UI toolkit.
Katherine Perry, Nathan Rizik, and Santiago Rodriguez Papa helped build software foundations for the experiments reported in this dissertation.
I am grateful for the opportunity to have worked with each of you.
On frustrating days with my own work, I am always glad to think of the good you're out to do.

In particular, Santiago Rodriguez Papa merits special recognition for jumping in the trenches to help bring this dissertation over the finish line.
Over the last four years, I have been endlessly entertained by your encyclopedic knowledge of amusing grotesqueries in society and technology.
I have also been endlessly impressed by your determination and cleverness in engineering better ways to do almost everything.
Thank you.

My own mentors invested time and personal support in my development.
Thank you to Dr. Rex Cole, Dr. America Chambers, Dr. John Fowler, Dr. Simon Garnier, Dr. Jason Graham, Mary Peterson, and Dr. Adam Smith.
Dr. America Chamber's encouraging and wise advisorship on my undergraduate thesis cemented my research interests and laid the foundation for my graduate career.
I am also grateful for my time training under the devoted mathematics and computer science faculty at the University of Puget Sound.

Thank you to Dr. Marisa Silver.
You are singularly responsible for getting me out the other end of middle school in one piece, with some serviceable writing ability to boot.

Without the benevolent grace of omnipotent administrative support staff, I could not have survived Michigan State University with a paycheck, health insurance, and graduation requirements.
Thank you to Barbara Bloemers, Deanne Hubbell, Connie James, and Melissa Williams.

This dissertation has benefited from the advice of my committee members: Dr. Wolfgang Banzhaf, Dr. Emily Dolson, Dr. Charles Ofria, and Dr. Bill Punch.
Thank you for especially for refining the focus of this work (and preventing global deforestation from production of a print copy).

What hasn't been said about Dr. Charles Ofria across dozens of advisorial acknowledgments?
I will add this: thank you most of all for planting yourself wholly in my corner.
From the very beginning, you made clear that I had your unconditional and total support.
I quickly grew to trust and rely on it.
I am glad to have been able to share my challenges with you, both technical and personal.
Thank you.

This research was supported in part by NSF grants DEB-1655715 and DBI-0939454 as well as by Michigan State University through the computational resources provided by the Institute for Cyber-Enabled Research.
This material is based upon work supported by the National Science Foundation Graduate Research Fellowship under Grant No. DGE-1424871.
Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the National Science Foundation.
